Modeling memory and taxis to analyze migrations

### 2021-05-14 text/ms.Rmd
working on ms

Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001].  Essentially, these models are grounded in the general idea that animal movements - somewhat like movements of physical particles - might combine a random (diffusive) component with a directed (advective) component and that the two can be separated and modeled. While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically. Even as equations analytically intractable,

Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g. by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics (**refs**).

@@

Seasonal migrations are an extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014]. For many species, it is an extremely successful strategy as it allows a greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996]. The proximate and ultimate causes and mechanisms, however, vary considerably across species, and even within species [@Shaw2016].Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015] and can be considered "tactical" migrants.  Others "jump" a green wave (or any other resource wave) anticipating that critical resources will be available at the time of arrival at the end point of migration (**refs**).  This second behavior involves a major trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized (**refs**).  In it's can be considered "strategic" in the sense that it is driven not by immediate cues but by memory-driven anticipation.  Clearly,

While migratory species are often considered to be more vulnerable to environmental change, as a disruptions in either of the seasonal ranges or along a migratory corridor can have significant negative impacts [@Wilcove2008;@Kauffman2021].  On the other hand, it has been argued that migratory species might be more resilient to dispruptions due to their wide-ranging mobility [@Robinson2009].

>>>>

Seasonal migrations are an extremely widespread among terrestrial, aquatic, avian and invertebrate species [@Dingle2014].  The ultimate evolutionary causes for migration are that the fitness benefits of accessing suitable seasonal resources, whether those are for energetic gain, predator avoidance, or suitably habitat for reproduction in respective seasons [@Avgar2013]. For many species, migration is an extremely successful strategy, allowing a far greater number of individuals to inhabit landscape which might not otherwise be able to support large numbers year round [@Fryxell1996].

Much as the ultimate causes of migration vary across species, the proximate causes, drivers and mechanism can vary considerably across and even within species [@Shaw2016]. Some migrants follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015] and can be considered "tactical" migrants.  Others "jump" a green wave (or any other resource wave) anticipating that critical resources will be available at the time of arrival at the end point of migration (**refs**).  This second behavior involves a major trade-off between the costs of migration against the benefits of accessing resources, whether food, suitable habitat for breeding, or predator refuge, that are highly seasonal and localized (**refs**).  This approach can be considered "strategic" in the sense that it is driven not by immediate cues but by memory-driven anticipation [@Bracis2017].

While migratory species are often considered to be more vulnerable to environmental change, as a disruption in either of the seasonal ranges or along a migratory corridor can have significant negative impacts [@Wilcove2008;@Kauffman2021].  On the other hand, it has been argued that migratory species might be more resilient to dispruptions due to their wide-ranging mobility [@Robinson2009]. Theoretical studies of migration indicate that some phenotypic plasticity is required to hedge both against persistent trends and increased stochasticity of resource distributions, but at the risk of sub-optimal access to resources [@Anderson2013].

Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001].  Essentially, these models are grounded in the general idea that animal movements - somewhat like movements of physical particles - might combine a random (diffusive) component with a directed (advective) component and that the two can be separated and modeled. While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically.Thus, much theoretical and some applied work has been done on refining the basic assumptions of diffusion models, e.g. by including heterogeneity in populations [@Skalski2003;@Gurarie2009], fat-tailed dispersal kernels [@Kot1996], non-linear or otherwise complex responses to resources and consepecifics (**refs**).

@@
